The Design and Implementation of the Synchronous Language \CEU

### 2016-03-28 paper.tex
...

% Title portion
\title{The Design, Semantics, and Implementation of the Synchronous Language 
\CEU}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward single-threaded implementation targeting resource-constrained 
systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C via function annotations and enforced 
finalization for external resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
% TODO: why? stack, fin, principalmente
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing that the peculiarities in the semantics of \CEU do not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, focusing on the fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

We decided for the unique and queue-based semantics for the reasons that 
follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since no time regularity is required for ticks~\cite{esterel.multiclock}, 
the two approaches for Esterel in Figure~\ref{fig.ticks} lead to different 
behaviors for the same sequence of inputs.
%
\item \textbf{Events are never simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but explicitly defined for each use case.
In the end of the section, we present a specification to detect 
``simultaneous'' button clicks.
%
Note that in the case of Esterel, simultaneity depends on the imprecise 
definition of ticks.
For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
simultaneous, even though \code{A} and \code{B} actually happen much closer to 
one another.
\item \textbf{Unique input events imply mutual exclusion:}
Because they are atomic, reactions to different events never overlap.
The possibility to reason about each input individually is a prerequisite for 
the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
simple expansions to a \code{par/and} and \code{par/or} for better readability.
%
Here, ``simultaneous'' means ``within 200 milliseconds'', which is a huge 
amount of time for a language-defined tick, and which would break the 
synchronous hypothesis.
%
A similar implementation for Esterel would not rely on the tick notion of 
simultaneity either.
%
We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
Section~\ref{sec.ceu.timers}.

TODO: recover the logical notion of simultaneaty substituting 200ms with TICK

@@

@@

@@

We presented the design, semantics, and implementation of \CEU, a synchronous 
reactive language inspired by Esterel targeting constrained embedded systems.

@@

% Idexx

>>>>

% Title portion
\title{The Design and Implementation of the Synchronous Language \CEU}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward single-threaded implementation targeting resource-constrained 
systems.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C via function annotations and enforced 
finalization for external resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, focusing on the fundamental 
differences to Esterel.
%Section~\ref{sec.sem} presents a formal semantics for the control primitives 
%of \CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

We decided for the unique and queue-based semantics for the reasons that 
follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but explicitly defined for each use case.
In the end of the section, we present a specification to detect 
``simultaneous'' button clicks.
%
Note that in the case of Esterel, simultaneity depends on the imprecise 
definition of ticks.
For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
simultaneous, even though \code{A} and \code{B} actually happen much closer to 
one another.
\item \textbf{Unique input events imply mutual exclusion:}
Because they are atomic, reactions to different events never overlap.
The possibility to reason about each input individually is a prerequisite for 
the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
Section~\ref{sec.ceu.timers}.
%
The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
simple expansions to a \code{par/and} and \code{par/or} for better readability.
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Note also that the application in \CEU could define an external input 
\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
10) to recover the abstract notion of simultaneity.

@@

lexical order pragamatics
    - again, familiar to programmers
        - register order
    - paralax, logging
        - simple, understandeable, imperative
        - multiple calls to compose, printf, are not commutative
        - define with intimidating name
    - implementation aspects in Section XX

@@

- understandable semantics
    - implementable

@@

We presented the design and implementation of \CEU, a synchronous reactive 
language inspired by Esterel targeting constrained embedded systems.

@@
